---
title: "biostats625"
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/xylin03/biostats625/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/xylin03/biostats625/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/xylin03/biostats625/branch/main/graph/badge.svg)](https://app.codecov.io/gh/xylin03/biostats625)
<!-- badges: end -->

## Overview

biostats625 provides a set of tools for performing ordinary least squares (OLS) linear regression in a simple and transparent way. Currently, it includes two main functions:

-   `fit_linear()` Fits a linear regression model after removing missing values. Returns coefficients, R-squared, adjusted R-squared, F-statistic, p-values, residuals, and fitted values.
-   `linreg_predict()` Predicts responses for new data using a model fitted by fit_linear().

These functions are designed to reproduce the behavior of base R’s `lm()` and `predict.lm()` while being simple enough for educational use. You can learn more about them in `vignette("linreg-vignette")`.

## Installation

```{r, eval = FALSE}
install.packages("devtools") # if you don't have devtools
devtools::install_github("xylin03/biostats625")
```

## Usage

```{r, message = FALSE}
# First, load the package
library(biostats625)

# Fit a linear regression model
# Example with mtcars dataset
model <- fit_linear(mpg ~ wt + hp + disp, data = mtcars)

# View coefficients
model$coefficients

# Check R-squared
model$r.squared

# Predict on new dataWriting
new_data <- data.frame(wt = c(2.5, 3.0), hp = c(110, 150), disp = c(109, 160))
preds <- predict_linear(model, new_data)
preds
```

## License

This package is released under the MIT License.
